### 2025-11-05 pca.qmd
20251101 - additions and clarifications
20251104 - typos and clarifications

As in [factor analysis](#sec-factorAnalysis).
The first decision when conducting a factor analysis is which variables to include and the scaling of those variables.
Guidance on which variables to include is in @sec-variablesToInclude-factorAnalysis.

@@

There are several differences between [factor analysis](#sec-factorAnalysis) and PCA.\index{factor analysis}\index{principal component analysis}
[Factor analysis](#sec-factorAnalysis) has greater sophistication than PCA, but greater sophistication often results in greater assumptions.\index{factor analysis}\index{principal component analysis}
[Factor analysis](#sec-factorAnalysis) does not always work; the data may not always fit to a [factor analysis](#sec-factorAnalysis) model.\index{factor analysis}\index{principal component analysis}
However, PCA can decompose any data matrix; it always works.\index{principal component analysis}
PCA is okay if you are not interested in the factor structure.\index{principal component analysis}
PCA uses all variance of variables and assumes variables have no error, so it does not account for measurement error.\index{principal component analysis}\index{measurement error}
PCA is good if you just want to form a linear composite and perform data reduction.\index{principal component analysis}\index{linear composite}\index{construct!formative}\index{construct!reflective}
However, if you are interested in the factor structure, use [factor analysis](#sec-factorAnalysis), which estimates a latent variable that accounts for the common variance and discards error variance.\index{factor analysis}
[Factor analysis](#sec-factorAnalysis) better handles error than PCA—[factor analysis](#sec-factorAnalysis) assumes that what is in the variable is the combination of common construct variance and error.\index{principal component analysis}\index{factor analysis}
By contrast, PCA assumes that the measures have no measurement error.\index{principal component analysis}\index{measurement error}
[Factor analysis](#sec-factorAnalysis) is useful for the identification of latent constructs—i.e., underlying dimensions or factors that explain (cause) observed scores.\index{factor analysis}

@@

The number of components to keep would depend on which criteria one uses.
Based on the rule to keep factors whose eigenvalues are greater than one and based on the parallel test, we would keep nine components.
However, based on the Cattell scree test (the "elbow" of the screen plot minus one), we would keep three components.
If using the optimal coordinates, we would keep four components; if using the acceleration factor, we would keep one component.
Therefore, interpretability of the components would be important for deciding how many components to keep.

>>>>

As in [factor analysis](#sec-factorAnalysis), the first decision when conducting a factor analysis is which variables to include and the scaling of those variables.
Guidance on which variables to include is in @sec-variablesToInclude-factorAnalysis.

@@

There are several differences between [factor analysis](#sec-factorAnalysis) and PCA.\index{factor analysis}\index{principal component analysis}
[Factor analysis](#sec-factorAnalysis) has greater sophistication than PCA, but greater sophistication often results in greater assumptions.\index{factor analysis}\index{principal component analysis}
[Factor analysis](#sec-factorAnalysis) does not always work; the data may not always fit to a [factor analysis](#sec-factorAnalysis) model.\index{factor analysis}\index{principal component analysis}
However, PCA can decompose any data matrix; it always works.\index{principal component analysis}
PCA is okay if you are not interested in the factor structure.\index{principal component analysis}
PCA uses all variance of variables and assumes variables have no error, so it does not account for measurement error.\index{principal component analysis}\index{measurement error}
PCA is good if you just want to form a linear composite and perform data reduction.\index{principal component analysis}\index{linear composite}\index{construct!formative}\index{construct!reflective}
However, if you are interested in the factor structure, use [factor analysis](#sec-factorAnalysis), which estimates a latent variable that accounts for the common variance and discards error variance.\index{factor analysis}
[Factor analysis](#sec-factorAnalysis) better handles error than PCA—[factor analysis](#sec-factorAnalysis) assumes that what is in the variable is the combination of common construct variance and error.\index{principal component analysis}\index{factor analysis}
By contrast, PCA uses the total variance (not just the common variance) and assumes that the measures have no measurement error.\index{principal component analysis}\index{measurement error}
[Factor analysis](#sec-factorAnalysis) is useful for the identification of latent constructs—i.e., underlying dimensions or factors that explain (cause) observed scores.\index{factor analysis}

@@

The number of components to keep would depend on which criteria one uses.
Based on the rule to keep factors whose eigenvalues are greater than one and based on the parallel test, we would keep nine components.
However, based on the Cattell scree test (the "elbow" of the screen plot minus one) [@Cattell1966], we would keep three components.
If using the optimal coordinates, we would keep four components; if using the acceleration factor, we would keep one component.
Therefore, interpretability of the components would be important for deciding how many components to keep.
